Energy Clustering

### 2017-10-12 energy_clustering/draft/energy_clustering.tex
aistats18

\begin{abstract}
Energy statistics was proposed by Sz\' ekely in the 80's inspired by the 
Newtonian gravitational potential from classical mechanics, and it provides 
a non-parametric hypothesis test for equality of distributions. 
It was further generalized from Euclidean spaces to metric spaces of 
strong negative type, and more recently, a connection with reproducing 
kernel Hilbert spaces (RKHS) was established. 
Here we consider the problem of clustering data from an 
energy statistics theory perspective.
We provide a precise mathematical formulation 
yielding a quadratically constrained 
quadratic program (QCQP) in the associated RKHS. We show that this QCQP
is equivalent to kernel $k$-means optimization problem, once the kernel
is fixed.
Thus, our results imply a first principles derivation of kernel $k$-means 
from energy statistics.
However, energy statistics fixes a family of standard kernels.
Furthermore, we also consider a weighted version of energy statistics, 
making connection to graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative algorithm based 
on Hartigan's method, which in this case has the same computational cost 
as kernel $k$-means algorithm, based on Lloyd's method, but usually 
with better clustering quality. 
We provide carefully designed numerical experiments showing the superiority 
of the proposed method compared to kernel $k$-means, standard $k$-means, 
and Gaussian mixture models in a variety of settings.
\end{abstract}

% @gui: you didn't update the abstract?

@@

% @gui: my experience is that most reviewers absolutely do *NOT* want a ToC, rather, we skip it, and journals often remove it prior to publication.

@@

Our work is organized as follows. In section~\ref{sec:background}, we introduce
the necessary background on energy statistics and RKHS.
Section~\ref{sec:clustering_theory} contains the main theoretical 
results of this paper,
where we consider a clustering theory based on energy statistics leading
to a QCQP, which is NP-hard in general. We also show the equivalence to 
kernel $k$-means optimization problem.
In Section~\ref{sec:weighted}, we generalize these results to a weighted
version of energy statistics, providing connections to graph
partitioning problems.
In Section~\ref{sec:twoclass}, we consider the  simple case of a two-class
problem in one dimension,
where we propose an algorithm which requires no random initialization.
In section~\ref{sec:algo}, we consider Lloyd's and Hartigan's methods
and propose iterative algorithms to solve the QCQP.
Section~\ref{sec:numerics} contains some carefully designed numerical
experiments showing that, in a variety of settings, energy clustering based
on Hartigan's algorithm
outperform $k$-means, GMM, kernel $k$-means, and a version of
spectral clustering.
Our final conclusions are presented in Section~\ref{sec:conclusion}.

@@

% @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
% @jovo: I stopped saying non-parametric, however, Kolmogorov-Smirnov is also
% said to be non-parametric. I think non-parametric tests also mean that
% it does not assume any form of distributions, i.e. it is a distribution-free
% statistical test.
% @gui: KS test is nonparametric because it can reject for any distribution, we cannot cluster for any distribution, so we are not.

@@

% @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
% @jovo: ok, will do that after we finish everything. There is a package that
% automatically remove numbers from unreferenced equations. It is a common
% practice however to number every equation in the paper,
% specially for reviewing purposes. But later 
% I will remove the numbers from the not cited ones if you don't like
% @gui: maybe common in math, but certainly not in stats / data science. 
% in any case, what is the package that does that?

% @gui: do other people know this result? (about min W = max X) if so, cite them, if not, state that.
% @jovo: eq (15), i.e. S + W = T is well-known. Gabor and Rizzo wrote that
% in their paper, however, they didn't write a proof. They should have
% it in some other paper.  However, they mention 
% that this holds only for 0 < alpha < 2, with |x-y|^alpha, while
% here I did for a general semimetric. Thus I don't really know if in general
% this is equality is known or not, but it should because it's a very easy
% derivation. The PhD thesis do not 
% mention anywhere that minimizing W is equivalent to maximizing S.
% Therefore, I don't want to sound like this is a big result and possibly
% a referee can say  "well this is trivial and known".
% Thus I just preferred to do my own derivation and not say anything.
% What do you think??
% @gui: i'd state it as a lemma then.

% @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
% @jovo: I tried to not commit to a single view. We show they are
% equivalent, when opperating on the same kernel. However, kernel k-means per
% se does not consider anything about the kernel choice. Energy statistics,
% however, provides a theory valid on spaces of negative type, which guarantees
% that a kernel constructed within this formalism will be 
% positive-semidefinite. Thus, I still prefer to look in the first direction.
% @gui: ok, did you update the text? i don't recall where i made this comment and don't want to search for it again. can you point me to the place that you say this?

@@

\begin{proposition}
\label{th:minimize}
Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
$x_i$ lives in a space $\mathcal{X}$ endowed with a semimetric $\rho:
\mathcal{X}\times\mathcal{X} \to \mathbb{R}$ of
negative type. For a fixed integer $k$,
the partition
$\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
all $i\ne j$, maximizes the between-sample statistic $S$, defined
in equation \eqref{eq:between}, if and only if
\begin{equation}
\label{eq:minimize}
\min_{\C_1,\dotsc,C_k  } W(
\C_1, \dotsc, \C_k),
\end{equation}
where the within energy dispersion $W$ is defined by \eqref{eq:within}.
\end{proposition}
\begin{proof}
From \eqref{eq:within} and \eqref{eq:between}
we have
\begin{equation}
\begin{split}
S + W &= 
\dfrac{1}{2n} \sum_{\substack{i,j=1 \\ i\ne j}}^k n_i n_j g(\C_i, \C_j)
+ \dfrac{1}{2n} \sum_{i=1}^{k} 
\bigg[ n - 
\sum_{j\ne i = 1}^k n_j \bigg] 
n_i g(\C_i, \C_i) \\
& = \dfrac{1}{2n} \sum_{i,j=1}^k n_i n_j g(\C_i, \C_j)
= \dfrac{1}{2n} \sum_{x \in \mathbb{X}} \sum_{y \in \mathbb{X}} \rho(x,y)
= \dfrac{n}{2} g(\mathbb{X}, \mathbb{X}).
\end{split}
\end{equation}
Note that the right hand side of this equation 
only depends on the pooled data, so it is a constant
independent of the choice of partition. Therefore, maximizing
$S$ over the choice of partition is equivalent to minimizing $W$.
\end{proof}

@@

\begin{proposition} 
\label{th:qcqp2}
The optimization problem \eqref{eq:minimize} is equivalent to
\begin{equation}
\label{eq:qcqp2}
\max_{\Zt} \Tr \left( \Zt^\top G \, \Zt \right)  \qquad
\mbox{s.t. $\Zt \ge 0$, $\Zt^\top \Zt = I$, 
$\Zt \Zt^\top \e = \e$},
\end{equation}
where $\e = (1,1,\dots,1)^\top \in \mathbb{R}^n$ is the all-ones vector,
and $G$ is the Gram matrix \eqref{eq:kernel_matrix}.
\end{proposition}
\begin{proof}
From 
\eqref{eq:gen_kernel},
\eqref{eq:g_def}, and
\eqref{eq:within}
we have
\begin{equation}
\label{eq:W2}
W(\C_1,\dotsc,\C_k  )
= \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y)
= \sum_{j=1}^k \sum_{x \in \C_j}  \bigg(
\kk(x,x) - \dfrac{1}{n_j} \sum_{y \in \C_j} \kk(x,y) \bigg).
\end{equation}
Note that the first term is global so it does not contribute to the 
optimization problem.
Therefore, minimizing \eqref{eq:W2} is equivalent to
\begin{equation}
\label{eq:max_prob}
\max_{ \C_1,\dotsc,\C_k } 
\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y\in C_j} \kk(x,y) .
\end{equation}
But 
\begin{equation}
\sum_{x, y \in \C_j} \kk(x, y) =
\sum_{p=1}^{n} \sum_{q=1}^{n} Z_{pj} Z_{qj} G_{pq} = 
(Z^\top G \, Z)_{jj},
\end{equation}
where we used the definitions \eqref{eq:kernel_matrix} and
\eqref{eq:label_matrix}. Thus, the objective function in 
\eqref{eq:max_prob} is equal to
$\Tr \left( D^{-1} Z^\top G Z \right)$. 
% @gui: the above step is not obvious
% @jovo: It is obvious. 
% From (21) we have \sum_j 1/n_j (Z^T G Z)_jj. But 1/n_j is just
% D^{-1}_jj (the diagonal matrix). The sum over j goes over the diagonals of
% D^{-1} Z^T G Z which is the definition of the trace
% @gui: obvious to you \neq obvious.  i understood, i want you to explain it.
Now we can
use the cyclic property
of the trace, and by the  definition of the matrix $Z$
in \eqref{eq:label_matrix}, we obtain the following integer
programing problem:
\begin{equation}\label{eq:qcqp}
\max_{Z} \Tr\Big( \big( Z D^{-1/2}\big)^\top G 
\big( ZD^{-1/2} \big) 
\Big) \quad
\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
$\sum_{i=1}^n Z_{ij} = n_j$.}
\end{equation}

Now we write this in terms of the matrix $Y = Z D^{-1/2}$.
The objective function immediately becomes
$\Tr\left( Y^\top G \, Y\right)$. Notice that the above constraints
imply that $Z^T Z = D$, where $D=\diag(n_1,\dotsc,n_k)$, which in turn gives
$D^{-1/2} Y^T Y D^{-1/2} = D$, or $Y^\top Y = I$. 
Also, every entry of $Y$ is positive by definition,
$Y \ge 0$. Now it only remains to show the last 
constraint in \eqref{eq:qcqp2}, which comes from the last
constraint in \eqref{eq:qcqp}. In matrix form this reads
$Z^T \e = D \e$. Replacing $Z=YD^{1/2}$ we have
$Y^\top \e = D^{1/2} \e$. Multiplying this last equation
on the left by $Y$, and noticing
that $Y D^{1/2} \e = Z \e = \e$, we finally obtain
$Y Y^\top \e = \e$. Therefore, the optimization 
problem \eqref{eq:qcqp} is equivalent
to \eqref{eq:qcqp2} .
\end{proof}

@@

The above result shows that 
kernel $k$-means problem is equivalent to the clustering problem
formulated in the energy statistics framework, when operating on the same
kernel. Notice, however, that 
energy statistics theory is valid for arbitrary semimetric spaces of
negative type, fixing the kernel function in the associated RKHS, which
is guaranteed to be positive definite.

@@

% @gui: "does not require any initiializaton" i think you mean is guaranteed to find the optimal solution in O(n^2) time.
% @jovo: it does not require initialization in the sense that we do not
% assign points to clusters as an initial step, as k-means++ for instance, or
% random initialization. This algorithm is "exact" since nothing
% is random in it, and multiple runs will always give the same answer
% @gui: i know what you mean, my point is that it is unclear.  state in the manuscript what you mean: "it is a deterministic algorithm." or state something else, that it does not require an initialization i don't feel is correct or informative.

@@

For a two-class problem we can use the formula
\eqref{eq:w1d} to cluster the data
through a simple algorithm 
as follows. We first order
the entire dataset, $\mathbb{X} \to \widetilde{\mathbb{X}}$. Then 
we compute \eqref{eq:w1d} for each possible split of $\widetilde{\mathbb{X}}$
and pick the point which gives the minimum value of $W$.
This procedure is described in Algorithm~\ref{algo1d} and called
$\mathcal{E}^{1D}$-clustering.
Note that this algorithm is exact in the sense that it does not involve
any random initialization,
however,
it only works for one-dimensional data with Euclidean distance. The total
complexity of $\mathcal{E}^{1D}$-clustering is $\OO(n\log n + n^2) = \OO(n^2)$.

@@

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=\textwidth]{1D_normal.pdf}\\[-1em](a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\[-.8em](b)
\end{minipage}
\caption{
\label{fig:1d}
$\mathcal{E}^{1D}$-clustering versus $k$-means and GMM in one dimension.
The two clusters have the same number of points, and we use
$50$ Monte Carlo runs.
We plot the average 
accuracy \eqref{eq:accuracy} versus the total number of points, 
where error bars are standard error. The dashed line indicates 
Bayes accuracy. (a) Data distributed as \eqref{eq:two_normal}, where the
optimal accuracy is $\approx 0.956$.
(b) Data from \eqref{eq:two_lognormal} where the optimal
accuracy is $\approx 0.852$.
% @gui: if you insist on using the equations, at least also mention something semantic about them, ie, Data distributed Gaussian (as in 46), or Data distributed log Gaussian (as in 47).  figures & captions should stand on their own without requiring referring to the paper.
}
% @gui: why did you take away the plot of the distribution? those are incredibly helpful.  
\end{figure}

@@

% @jovo: check this paragraph. I included it to answer your question
% about having K(x,x) and \ell=1,...,k without exluding \ell=j. Essentially,
% the term K(x,x) will be hidden inside the mean of the cluster to which
% x is currently assigned to. This also clarifies the connection to 
% the standard formulation of kernel k-means.
$\mathcal{E}^{L}$-clustering is precisely
kernel $k$-means algorithm \cite{Dhillon2,Dhillon} 
but written more concisely and with the kernel induced by energy statistics. Indeed, recalling that
$\kk(x,y)=\langle \varphi(x), \varphi(y) \rangle$ where $\varphi: \mathcal{X}
\to \mathcal{H}_\kk$ is the feature map, we have from \eqref{eq:Jell} that
\begin{equation}
%\begin{split}
J^{(\ell)}(x_i) 
%&= \dfrac{1}{n_\ell^2} \sum_{x, y \in \C_\ell} 
%\langle \varphi(x), \varphi(y) \rangle - \dfrac{2}{n_\ell}
%\sum_{y\in \C_\ell} \langle \varphi(x_i), \varphi(y)\rangle \\ 
= 
\langle \varphi(\mu_\ell), \varphi(\mu_\ell) \rangle
-2 \langle \varphi(x_i), \varphi(\mu_\ell) \rangle 
= 
\| \varphi(x_i) - \varphi(\mu_\ell) \|^2 - \| \varphi(x_i) \|^2,
%\end{split}
\end{equation}
where $\mu_\ell = \tfrac{1}{n_\ell} \sum_{x\in \C_\ell}x$ 
is the mean of cluster $\C_\ell$. 
Therefore, $\min_\ell J^{(\ell)}(x_i) = 
\min_\ell \| \varphi(x_i) - \varphi(\mu_\ell)\|^2$, i.e. we are assigning
$x_i$ to the cluster with closest center (in feature space),
which is the familiar Lloyd's
heuristic approach that kernel $k$-means is based upon.

@@

\begin{figure}
\begin{flushleft}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
    \INPUT number of clusters $k$, Gram matrix $G$, initial label
    matrix $Z \leftarrow Z_0$
    \OUTPUT label matrix $Z$ 
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            have the costs of each cluster, defined in \eqref{eq:maxQ}
  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ 
        have the number of points in each cluster%, obtained 
        %from $D = Z^\top Z$
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i)$,
            where $J^{(\ell)}(x_i)$ is defined in \eqref{eq:Jell}
            %, for $\ell=1,2,\dots,k$
        \IF{ $j^\star \ne j$} 
            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and
            $Z_{ij^\star} \leftarrow 1$
            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{kmeans_algo}
$\mathcal{E}^{L}$-clustering is Lloyd's method for energy clustering, which
is precisely~~~~~~~~ \\ 
kernel $k$-means algorithm, with the kernel induced by energy statistics. This procedure finds
local solutions to the problem \eqref{eq:qcqp2}.\hspace{\fill}
% @gui: spacing is weird.
}
\end{algorithm}
\end{flushleft}
\end{figure}

@@

% @jovo: I also clarified this section, specially the derivation
% of the algorithm. Actually there was a mistake in the previous formula,
% which carried over to my code. Now it's fixed and better.

@@

@@

% @gui: something is weird with the way you care computing accuracy.  GMM should not be significantly worse than chance.  i think you are probably counting the NaN's as *wrong*, rather than nothing?
% now I'm using GMM from sklearn which eliminated previous weird things.
% I think my method was too crude and the covariance estimate was too bad in
% some cases

% @gui: i'm not convinced you need alg 2 in these tables at all. if i were to add alg 2 to these tables, however, i would do the same thing i would for the figs, that is, for each simulation, compute the diff between performance on alg 2 and alg 3, and report that.
% @jovo: yes I agree. I included new plots in the fig, but didn't do anything
% for the tables cuz I don't think necessary

>>>>

\begin{abstract}
Energy statistics was proposed by Sz\' ekely in the 80's inspired by the 
Newtonian gravitational potential from classical mechanics, and it provides 
a 
hypothesis test for equality of distributions. 
It was further generalized from Euclidean spaces to metric spaces of 
strong negative type, and more recently, a connection with reproducing 
kernel Hilbert spaces (RKHS) was established. 
Here we consider the problem of clustering data from an 
energy statistics theory perspective.
We provide a precise mathematical formulation 
yielding a quadratically constrained 
quadratic program (QCQP) in the associated RKHS. We show that this QCQP
is equivalent to kernel $k$-means optimization problem, once the kernel
is fixed.
Thus, our results imply a first principles derivation of kernel $k$-means 
from energy statistics.
However, energy statistics fixes a family of standard kernels.
Furthermore, we also consider a weighted version of energy statistics, 
making connection to graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative algorithm based 
on Hartigan's method, which in this case has the same computational cost 
as kernel $k$-means algorithm, based on Lloyd's method, but usually 
with better clustering quality. 
We provide carefully designed numerical experiments showing the superiority 
of the proposed method compared to kernel $k$-means, standard $k$-means, 
and Gaussian mixture models in a variety of settings.
\end{abstract}

% @gui: you didn't update the abstract?
% @jovo: I did, removed some stuff that you didn't like before, now
% removing the non-parametric word. You think it's not good enough?

@@

@@

%Our work is organized as follows. In section~\ref{sec:background}, we introduce
%the necessary background on energy statistics and RKHS.
%Section~\ref{sec:clustering_theory} contains the main theoretical 
%results of this paper,
%where we consider a clustering theory based on energy statistics leading
%to a QCQP, which is NP-hard in general. We also show the equivalence to 
%kernel $k$-means optimization problem.
%In Section~\ref{sec:weighted}, we generalize these results to a weighted
%version of energy statistics, providing connections to graph
%partitioning problems.
%In Section~\ref{sec:twoclass}, we consider the  simple case of a two-class
%problem in one dimension,
%where we propose an algorithm which requires no random initialization.
%In section~\ref{sec:algo}, we consider Lloyd's and Hartigan's methods
%and propose iterative algorithms to solve the QCQP.
%Section~\ref{sec:numerics} contains some carefully designed numerical
%experiments showing that, in a variety of settings, energy clustering based
%on Hartigan's algorithm
%outperform $k$-means, GMM, kernel $k$-means, and a version of
%spectral clustering.
%Our final conclusions are presented in Section~\ref{sec:conclusion}.

@@

@@

% @gui: maybe common in math, but certainly not in stats / data science. 
% in any case, what is the package that does that?
% @jovo: \usepackage{mathtools} \mathtoolsset{showonlyrefs}

% @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
% @jovo: I'll put a comment below, I simply removed the previous phrase,
% didn't say anything instead.

@@

\begin{lemma}
\label{th:minimize}
Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
$x_i$ lives in a space $\mathcal{X}$ endowed with a semimetric $\rho:
\mathcal{X}\times\mathcal{X} \to \mathbb{R}$ of
negative type. For a fixed integer $k$,
the partition
$\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
all $i\ne j$, maximizes the between-sample statistic $S$, defined
in equation \eqref{eq:between}, if and only if
\begin{equation}
\label{eq:minimize}
\min_{\C_1,\dotsc,C_k  } W(
\C_1, \dotsc, \C_k),
\end{equation}
where the within energy dispersion $W$ is defined by \eqref{eq:within}.
\end{lemma}
\begin{proof}
From \eqref{eq:within} and \eqref{eq:between}
we have
\begin{equation}
\begin{split}
S + W &= 
\dfrac{1}{2n} \sum_{\substack{i,j=1 \\ i\ne j}}^k n_i n_j g(\C_i, \C_j)
+ \dfrac{1}{2n} \sum_{i=1}^{k} 
\bigg[ n - 
\sum_{j\ne i = 1}^k n_j \bigg] 
n_i g(\C_i, \C_i) \\
& = \dfrac{1}{2n} \sum_{i,j=1}^k n_i n_j g(\C_i, \C_j)
= \dfrac{1}{2n} \sum_{x \in \mathbb{X}} \sum_{y \in \mathbb{X}} \rho(x,y)
= \dfrac{n}{2} g(\mathbb{X}, \mathbb{X}).
\end{split}
\end{equation}
Note that the right hand side of this equation 
only depends on the pooled data, so it is a constant
independent of the choice of partition. Therefore, maximizing
$S$ over the choice of partition is equivalent to minimizing $W$.
\end{proof}

@@

\begin{proposition} 
\label{th:qcqp2}
The optimization problem \eqref{eq:minimize} is equivalent to
\begin{equation}
\label{eq:qcqp2}
\max_{\Zt} \Tr \left( \Zt^\top G \, \Zt \right)  \qquad
\mbox{s.t. $\Zt \ge 0$, $\Zt^\top \Zt = I$, 
$\Zt \Zt^\top \e = \e$},
\end{equation}
where $\e = (1,1,\dots,1)^\top \in \mathbb{R}^n$ is the all-ones vector,
and $G$ is the Gram matrix \eqref{eq:kernel_matrix}.
\end{proposition}
\begin{proof}
From 
\eqref{eq:gen_kernel},
\eqref{eq:g_def}, and
\eqref{eq:within}
we have
\begin{equation}
\label{eq:W2}
W(\C_1,\dotsc,\C_k  )
= \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y)
= \sum_{j=1}^k \sum_{x \in \C_j}  \bigg(
\kk(x,x) - \dfrac{1}{n_j} \sum_{y \in \C_j} \kk(x,y) \bigg).
\end{equation}
Note that the first term is global so it does not contribute to the 
optimization problem.
Therefore, minimizing \eqref{eq:W2} is equivalent to
\begin{equation}
\label{eq:max_prob}
\max_{ \C_1,\dotsc,\C_k } 
\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y\in C_j} \kk(x,y) .
\end{equation}
But 
\begin{equation}
\sum_{x, y \in \C_j} \kk(x, y) =
\sum_{p=1}^{n} \sum_{q=1}^{n} Z_{pj} Z_{qj} G_{pq} = 
(Z^\top G \, Z)_{jj},
\end{equation}
where we used the definitions \eqref{eq:kernel_matrix} and
\eqref{eq:label_matrix}. 
Notice that $n_j^{-1} = D^{-1}_{jj}$, where the diagonal matrix $D = 
\diag(n_1,\dotsc,n_k)$ contains the number of points in each cluster, 
thus the objective function in 
\eqref{eq:max_prob} is equal to $\sum_{j=1}^k D^{-1}_{jj} 
\left( Z^\top G Z \right)_{jj} = \Tr \left( D^{-1} Z^\top G Z \right)$. 
Now we can
use the cyclic property
of the trace, and by the  definition of the matrix $Z$
in \eqref{eq:label_matrix}, we obtain the following integer
programing problem:
\begin{equation}\label{eq:qcqp}
\max_{Z} \Tr\Big( \big( Z D^{-1/2}\big)^\top G 
\big( ZD^{-1/2} \big) 
\Big) \quad
\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
$\sum_{i=1}^n Z_{ij} = n_j$.}
\end{equation}

Now we write this in terms of the matrix $Y = Z D^{-1/2}$.
The objective function immediately becomes
$\Tr\left( Y^\top G \, Y\right)$. Notice that the above constraints
imply that $Z^T Z = D$, which in turn gives
$D^{-1/2} Y^T Y D^{-1/2} = D$, or $Y^\top Y = I$. 
Also, every entry of $Y$ is positive by definition,
$Y \ge 0$. Now it only remains to show the last 
constraint in \eqref{eq:qcqp2}, which comes from the last
constraint in \eqref{eq:qcqp}. In matrix form this reads
$Z^T \e = D \e$. Replacing $Z=YD^{1/2}$ we have
$Y^\top \e = D^{1/2} \e$. Multiplying this last equation
on the left by $Y$, and noticing
that $Y D^{1/2} \e = Z \e = \e$, we finally obtain
$Y Y^\top \e = \e$. Therefore, the optimization 
problem \eqref{eq:qcqp} is equivalent
to \eqref{eq:qcqp2} .
\end{proof}

@@

% @jovo: the previous paragraph saying that kernel k-means is a consequence
% of energy statistics WAS here, I removed and now don't say anything.
The above result shows that 
kernel $k$-means problem is equivalent to the clustering problem
formulated in the energy statistics framework, when operating on the same
kernel. Notice, however, that 
energy statistics theory is valid for arbitrary semimetric spaces of
negative type, fixing the kernel function in the associated RKHS, which
is guaranteed to be positive definite.

@@

@@

For a two-class problem we can use the formula
\eqref{eq:w1d} to cluster the data
through a simple algorithm 
as follows. We first order
the entire dataset, $\mathbb{X} \to \widetilde{\mathbb{X}}$. Then 
we compute \eqref{eq:w1d} for each possible split of $\widetilde{\mathbb{X}}$
and pick the point which gives the minimum value of $W$.
This procedure is described in Algorithm~\ref{algo1d} and called
$\mathcal{E}^{1D}$-clustering.
Note that this algorithm is deterministic,
however,
it only works for one-dimensional data with Euclidean distance. The total
complexity of $\mathcal{E}^{1D}$-clustering is $\OO(n\log n + n^2) = \OO(n^2)$.

@@

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=\textwidth]{1D_normal.pdf}\\[-1em](a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\[-.8em](b)
\end{minipage}
\caption{
\label{fig:1d}
$\mathcal{E}^{1D}$-clustering versus $k$-means and GMM in one dimension.
The two clusters have the same number of points, and we use
$50$ Monte Carlo runs.
We plot the average 
accuracy \eqref{eq:accuracy} versus the total number of points, 
where error bars are standard error. The dashed line indicates 
Bayes accuracy. (a) Data normally distributed as in 
\eqref{eq:two_normal}, where the
optimal accuracy is $\approx 0.956$.
(b) Data log-normally distributed as in \eqref{eq:two_lognormal} where 
the optimal accuracy is $\approx 0.852$.
}
% @gui: if you insist on using the equations, at least also mention something semantic about them, ie, Data distributed Gaussian (as in 46), or Data distributed log Gaussian (as in 47).  figures & captions should stand on their own without requiring referring to the paper.
% @gui: why did you take away the plot of the distribution? those are incredibly helpful.
% @jovo: ok will restore them.
\end{figure}

@@

$\mathcal{E}^{L}$-clustering is precisely
kernel $k$-means algorithm \cite{Dhillon2,Dhillon} 
but written more concisely and with the kernel induced by energy statistics. 
Indeed, recalling that
$\kk(x,y)=\langle \varphi(x), \varphi(y) \rangle$ where $\varphi: \mathcal{X}
\to \mathcal{H}_\kk$ is the feature map, we have from \eqref{eq:Jell} that
\begin{equation}
%\begin{split}
J^{(\ell)}(x_i) 
%&= \dfrac{1}{n_\ell^2} \sum_{x, y \in \C_\ell} 
%\langle \varphi(x), \varphi(y) \rangle - \dfrac{2}{n_\ell}
%\sum_{y\in \C_\ell} \langle \varphi(x_i), \varphi(y)\rangle \\ 
= 
\langle \varphi(\mu_\ell), \varphi(\mu_\ell) \rangle
-2 \langle \varphi(x_i), \varphi(\mu_\ell) \rangle 
= 
\| \varphi(x_i) - \varphi(\mu_\ell) \|^2 - \| \varphi(x_i) \|^2,
%\end{split}
\end{equation}
where $\mu_\ell = \tfrac{1}{n_\ell} \sum_{x\in \C_\ell}x$ 
is the mean of cluster $\C_\ell$. 
Therefore, $\min_\ell J^{(\ell)}(x_i) = 
\min_\ell \| \varphi(x_i) - \varphi(\mu_\ell)\|^2$, i.e. we are assigning
$x_i$ to the cluster with closest center (in feature space),
which is the familiar Lloyd's
heuristic approach that kernel $k$-means is based upon.

@@

\begin{figure}
\begin{flushleft}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
    \INPUT number of clusters $k$, Gram matrix $G$, initial label
    matrix $Z \leftarrow Z_0$
    \OUTPUT label matrix $Z$ 
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            have the costs of each cluster, defined in \eqref{eq:maxQ}
  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ 
        have the number of points in each cluster%, obtained 
        %from $D = Z^\top Z$
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i)$,
            where $J^{(\ell)}(x_i)$ is defined in \eqref{eq:Jell}
            %, for $\ell=1,2,\dots,k$
        \IF{ $j^\star \ne j$} 
            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and
            $Z_{ij^\star} \leftarrow 1$
            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{kmeans_algo}
$\mathcal{E}^{L}$-clustering is Lloyd's method for energy clustering, which
is precisely~~~~~~~~~ \\
kernel $k$-means algorithm, with the kernel induced by energy statistics. 
This procedure~~~~~~~ \\
finds
local solutions to the problem \eqref{eq:qcqp2}.\hspace{\fill}
}
\end{algorithm}
\end{flushleft}
\end{figure}

@@

@@

\begin{figure}
\begin{flushleft}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
    \INPUT number of clusters $k$, Gram matrix $G$, 
                initial label matrix $Z \leftarrow Z_0$
    \OUTPUT label matrix $Z$
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            have the energy costs of each cluster, defined in \eqref{eq:maxQ}
  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ have the number of points 
        in each cluster%, obtained from $D=Z^\top Z$
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
                \Delta Q^{j\to \ell}(x_i)$
            %for $\ell=1,2,\dots,k$ and $\ell \ne j$; 
            using \eqref{eq:changeQ} \label{stepmove}
        \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and 
            $Z_{ij^\star} \leftarrow 1$
            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i) + G_{ii}$ and
    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)+ G_{ii}$
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{algo}
$\mathcal{E}^H$-clustering is Hartigan's method for energy clustering.
This algorithm~~~~~~ \\ finds local solutions to  
the optimization problem \eqref{eq:qcqp2}.
The steps $6$ and $10$ are different~~~~~~~~ \\
than $\mathcal{E}^L$-clustering
described in Algorithm~\ref{kmeans_algo}. \hspace{\fill}
}
\end{algorithm}
\end{flushleft}
\end{figure}

@@

The main goal of this section is threefold. First, we want to compare
energy clustering, based on the kernels provided by the standard energy
distance $\mathcal{E}_\alpha$, to $k$-means and GMM. Second, we want
to compare $\mathcal{E}^H$-clustering, based on Hartigan's method, to
$\mathcal{E}^L$-clustering, based on Lloyd's method. We also compare
to $\mathcal{E}^S$-clustering
which is an exact solution to the relaxed problem.
Third, we want to illustrate the flexibility
provided by energy clustering which do not make assumptions about the
distribution of the data, being able to cluster accurately in different
settings while keeping the same standard kernel.
